The forestecology R package for fitting and assessing models of interspecies competitive effects on the growth of trees

### 2021-02-23 paper/forestecology.Rmd
Here are my edits. A couple of things - This is great! Very nice clear explanation of each step to do this analysis. - I am not 100% sure, but MEE might want a little bit more on why the underlying method is important. I could be wrong. Maybe the PLOS One paper serves that purpose and all we are doing here is presenting the package and explaning how to use it. - With the above said, I think what you have here is great and no reason to do a major restructure. But just make a clear push in the intro for the underlying method. - I have a couple of places hwere I hvae left commetns/notes. YOu can

Repeat-censused forest plots offer excellent data to test neighborhood models of tree competition @allen_permutation_2020 @canham_neighborhood_2006 @uriarte_spatially_2004. Here we describe an R package, `forestecology`, to do that. This package implements the methods in @allen_permutation_2020. It provides: a convenient way to specify and fit models of tree growth based on neighborhood competition; a spatial cross validation method to test and compare model fits @roberts_cross-validation_2017; and an ANOVA-like method to assess whether the competitor identity matters in these models. The model is written to work with ForestGEO plot data @andersonteixeira_ctfs-forestgeo_2015, but we envision that it could easily be modified to work with data from other forest plots, e.g. the US Forest Service Forest Inventory and Analysis plots @smith_forest_2002.

@@

We demonstrate the `forestecology` package's functionality on data from the Smithsonian Conservation Biology Institute (SCBI) large forest dynamics plot, located at the Smithsonian's National Zoo and Conservation Biology Institute in Front Royal, VA, USA. The 25.6 ha (640 x 400 m) plot is located at the intersection of three of the major physiographic provinces of the eastern US: the Blue Ridge, Ridge and Valley, and Piedmont provinces and is adjacent to the northern end of Shenandoah National Park. The forest type is typical mature secondary eastern mixed deciduous forest, with a canopy dominated by tulip poplar (*Liriodendron tulipifera*), oaks (*Quercus* spp.), and hickories (*Carya* spp.), and an understory composed mainly of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*) @bourg_initial_2013.

The `forestecology` package has the following ecological goals: 1) to evaluate the effect of competitor species identity using permutation tests and 2) to evaluate model performance using spatial cross-validation. To achieve these goals, we outline a basic analysis sequence comprising of these four main steps:

1. Compute the growth of trees based on census data.
1. Add spatial information:
    1. Define a buffer region of trees.
    1. Add spatial cross-validation block information.
1. Identify all focal trees and their competitors.
1. Apply model, which includes:
    1. Fit model.
    1. Compute fitted/predicted values.
    1. Visualize posterior distributions.

@@

The first step in the our analysis sequence is to compute the growth of trees using data from two censuses. The `compute_growth()` function computes growth assuming census data that roughly follows ForestGEO standards. Despite such standards, minor variations will still exist between sites, thereby necessitating some data wrangling and checking. For example, the SCBI site records all DBH's in millimeters @bourg_initial_2013, whereas the Michigan Big Woods site records them in centimeters @allen_michigan_2020.

We load both 2008 and 2014 SCBI census data `.csv` files as they existed on GitHub on November 20, 2020. After selecting only the relevant variables, we perform a few additional data wrangling steps: convert the character variable with the date of measurement to be of explicit type `date`, convert DBH to be in centimeters^[A rule of thumb to determine the units of DBH is to check if the smallest non-zero and non-missing measurement is 1 or 10. If the former, then centimeters. If the later, then millimeters. This is because ForestGEO protocols state that only trees with DBH greater or equal to 1cm should be included in censuses.], convert the `sp` variable containing species information from type `chr` character to `fct` factor.^[In our spatial cross-validation algorithm in Section \ref{spatial-cross-validation} issues can occur when rare species do not occur in the training set, but then are encountered in the test set. This risk is mitigated by representing `sp` as a factor variable, which has a complete list of all levels of the categorical variable.]Furthermore, in order to speed up computation for purposes of this example, we only consider a 9 ha subsection of the 25.6 ha of the SCBI site: `gx` from 0--300 instead of 0--400 and `gy` from 300--600 instead of 0--640.

@@

These two data frames are then used as arguments to the `compute_growth()` function, along with the `id` argument that specifies the variable that uniquely identifies each tree-stem. Note furthermore that we discard all resprouts in the later census (those with `code == R`).

@@

The output `growth_scbi` is a single data frame of class `sf` that includes `growth`, the average annual growth in DBH (in cm) for all trees that were alive at both time points, as well a `geometry` variable encoding each tree's geolocation. Furthermore, the variables that should remain unchanged between censuses appear only once, such as location variables `gx` and `gy`; as well as species-related variables. Variables that should change between censuses are suffixed with `1` and `2` indicating the earlier and later censuses, such as `dbh1/dbh2` and `codes1/codes2`.

The data format of other sites may be such that our `compute_growth()` function doesn't work at all. However, in the end all that matters is that the growth of all trees is saved in a data frame of class `sf` whereby the geolocation of each tree is presented in a `geometry` variable of type `<POINT>` and at a minimum the data contains the variables above.

@@

The next step in our analysis sequence is to add spatial information to our main `growth_scbi` data frame. The first element of spatial information we add is a "buffer region" to the periphery of the study region. Since some of our model's explanatory variables are cumulative (such as competitor basal area), we must ensure that all trees being modeled are not biased to have different neighbor structures. This is of concern for trees at the boundary of study regions, which will not have the same number of neighbors as trees in the internal part of the study region. In order to account for such edge effects, only trees that are not part of this buffer region, i.e. are part of the interior of the study region, will have their growths modeled @waller_applied_2004.

Our model of interspecific competition relies on a spatial definition of who the competitor trees are for focal trees of interest: all trees within a distance `comp_dist` of a focal tree are considered its competitors (assuming the same units as the `gx` and `gy` location variables). In our case we set this value at 7.5m, a value informed by @canham_neighborhood_2004 @uriarte_spatially_2004 @canham_neighborhood_2006. Using this value along with a manually constructed `sf` object representation of the study region's boundary, we apply the `add_buffer_variable()` to our `growth_scbi` data frame to add a `buffer` boolean variable: all trees who have `buffer` set to `FALSE` will be our focal trees whose growths are modeled, whereas those with `TRUE` will only be considered as competitor trees whose growth will not be modeled.

@@

The second element of spatial information are blocks corresponding to folds of a spatial cross-validation algorithm used to estimate out-of-sample model error. Conventional cross-validation algorithms assign observations to folds by randomly resampling individual observations. However, an assumption that the observations are independent underlies any cross-validation algorithm. In the case of forest census data, observations exhibit spatial autocorrelation. We therefore incorporate this spatial dependence into the cross-validation algorithm by randomly resampling spatial blocks of trees @roberts_cross-validation_2017 @pohjankukka_estimating_2017.

@@

@@

We then compare the observed and fitted/predicted growths to compute the root mean squared error (RMSE) of our model fit.

```{r scbi-model-rmse}
model_rmse <- focal_vs_comp_scbi %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
model_rmse
```

@@

TODO: Discuss meaning of results.

@@

model_rmse_shuffle <- focal_vs_comp_scbi %>%
  rmse(truth = growth, estimate = growth_hat_shuffle) %>%
  pull(.estimate)
model_rmse_shuffle
```

The resulting RMSE of `r model_rmse_shuffle` based on the permutation test is `r ifelse(model_rmse_shuffle < model_rmse, "smaller", "larger")` than the earlier RMSE of `r model_rmse`, suggesting that models that do incorporate competitor species identity better fit the data. We conduct a fuller simulation in Section below.

@@

To this end, we use the `foldID` variable defined in Section \ref{spatial-information} whereby all focal trees are assigned to one of `r n_fold` spatially contiguous blocks that act as folds in our cross-validation routine. Figure \ref{fig:scbi-spatial-cross-validation-schematic} presents a schematic illustrating this scheme for fold 1 (bottom-left) as the test set and folds 2, 3, and 4 as the training sets. We fit the model to all focal trees in the training set, apply the model to all focal trees in the test set to compute fitted/predicted values, and compute the RMSE of the observed versus predicted growths. We repeat this procedure 3 more times with each of the three remaining folds acting as the test set and then average all four resulting RMSE's. Furthermore, in order to maintain spatial independence between the test and training set, a buffer that extend outwards from the boundary of the test set is computed; all trees falling within this buffer are excluded from the training set.

```{r scbi-spatial-cross-validation-schematic, out.width="66%", fig.cap= "Schematic of spatial cross-validation: Using the k = 1 fold as the test set, assigning each focal tree to training set, test set, and buffer.", echo = FALSE}
fold_number <- 1

@@

test_vs_train <- focal_vs_comp_scbi %>% 
  st_as_sf() %>% 
  add_buffer_variable(
    direction = "out", size = comp_dist, region = test_fold_buffer
  ) %>% 
  mutate(
    fold = case_when(
      foldID == fold_number ~ "test",
      foldID != fold_number & buffer ~ "training",
      foldID != fold_number & !buffer ~ "buffer"
    ),
    fold = factor(fold, levels = c("test", "buffer", "training"))
  )

@@

```{r scbi-spatial-cv}
focal_vs_comp_scbi <- focal_vs_comp_scbi %>%
  run_cv(comp_dist = comp_dist, cv_grid = blocks_scbi)
```
```{r}
model_rmse_cv <- focal_vs_comp_scbi %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
model_rmse_cv
```

>>>>

Repeat-censused forest plots offer excellent data to test neighborhood models of tree competition @allen_permutation_2020 @canham_neighborhood_2006 @uriarte_spatially_2004. Here we describe an R package, `forestecology`, to do that. This package implements the methods in @allen_permutation_2020. It provides: a convenient way to specify and fit models of tree growth based on neighborhood competition; a spatial cross validation method to test and compare model fits @roberts_cross-validation_2017; and an ANOVA-like method to assess whether the competitor identity matters in these models. The model is written to work with ForestGEO plot data @andersonteixeira_ctfs-forestgeo_2015, but we envision that it could easily be modified to work with data from other forest plots, e.g. the US Forest Service Forest Inventory and Analysis plots @smith_forest_2002. Or more generally to model interactions of any community of mapped sessile organisms.

@@

We demonstrate the `forestecology` package's functionality on data from the Smithsonian Conservation Biology Institute (SCBI) large forest dynamics plot, located at the Smithsonian's National Zoo and Conservation Biology Institute in Front Royal, VA, USA @bourg_initial_2013. The 25.6 ha (640 x 400 m) plot is located at the intersection of three of the major physiographic provinces of the eastern US: the Blue Ridge, Ridge and Valley, and Piedmont provinces and is adjacent to the northern end of Shenandoah National Park. The forest type is typical mature secondary eastern mixed deciduous forest, with a canopy dominated by tulip poplar (*Liriodendron tulipifera*), oaks (*Quercus* spp.), and hickories (*Carya* spp.), and an understory composed mainly of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*) @bourg_initial_2013.

The `forestecology` package has the following ecological goals: 1) to evaluate the effect of competitor species identity on using permutation tests and 2) to evaluate model performance using spatial cross-validation. To achieve these goals, we outline a basic analysis sequence comprising of these four main steps:

1. Compute the growth of stems based on two censuses.
1. Add spatial information:
    1. Define a buffer region of trees.
    1. Add spatial cross-validation block information.
1. Identify all focal trees and their competitors.
1. Apply model, which includes:
    1. Fit model.
    1. Compute fitted/predicted values.
    1. Visualize posterior distributions.

@@

The first step in the our analysis sequence is to compute the growth of trees using data from two censuses. The `compute_growth()` function computes average annual growth assuming census data that roughly follows ForestGEO standards. Despite such standards, minor variations will still exist between sites, thereby necessitating some data wrangling and checking. For example, the SCBI site records all diameters at breast height (DBH) in millimeters @bourg_initial_2013, whereas the Michigan Big Woods site records them in centimeters @allen_michigan_2020.

We load both 2008 and 2014 SCBI census data `.csv` files as they existed on GitHub on November 20, 2020. After selecting only the relevant variables, we perform a few additional data wrangling steps: convert the character variable with the date of measurement to be of explicit type `date`, convert DBH to be in centimeters^[A rule of thumb to determine the units of DBH is to check if the smallest non-zero and non-missing measurement is 1 or 10. If the former, then centimeters. If the later, then millimeters. This is because ForestGEO protocols state that only trees with DBH greater or equal to 1cm should be included in censuses.], convert the `sp` variable containing species information from type `chr` character to `fct` factor.^[In our spatial cross-validation algorithm in Section \ref{spatial-cross-validation} issues can occur when rare species do not occur in the training set, but then are encountered in the test set. This risk is mitigated by representing `sp` as a factor variable, which has a complete list of all levels of the categorical variable.] Furthermore, in order to speed up computation for purposes of this example, we only consider a 9 ha subsection of the 25.6 ha of the SCBI site: `gx` from 0--300 instead of 0--400 and `gy` from 300--600 instead of 0--640.

@@

These two data frames are then used as arguments to the `compute_growth()` function, along with the `id` argument that specifies the variable that uniquely identifies each tree-stem. Note furthermore that we discard all resprouts in the later census (those with `code == R`), since we are only interested in the diameter growth of surviving, and not resprouted, stems.

@@

The output `growth_scbi` is a single data frame of class `sf` that includes `growth`, the average annual growth in DBH (cm y^-1^) for all stems that were alive at both time points, as well a `geometry` variable encoding each tree's geolocation. Furthermore, the variables that should remain unchanged between censuses appear only once, such as location variables `gx` and `gy`; as well as species-related variables. Variables that should change between censuses are suffixed with `1` and `2` indicating the earlier and later censuses, such as `dbh1/dbh2` and `codes1/codes2`.

The data format of other sites may be such that our `compute_growth()` function doesn't work at all. However, in the end all that matters is that the growth of all stems is saved in a data frame of class `sf` whereby the geolocation of each tree is presented in a `geometry` variable of type `<POINT>` and at a minimum the data contains the variables above (DAVE NOTE: THis isn't exaclty clear to me what variables above means. Maybe list all the variables needed. Do they have to have exact names?)

@@

The next step in our analysis sequence is to add additional spatial information to our main `growth_scbi` data frame. The first element of spatial information we add is a "buffer region" to the periphery of the study region. Since some of our model's explanatory variables are cumulative (such as competitor basal area), we must ensure that all trees being modeled are not biased to have different neighbor structures. This is of concern for trees at the boundary of study regions, for which all neighbors will not be included in the censused stems. In order to account for such edge effects, only trees that are not part of this buffer region, i.e. are part of the interior of the study region, will have their growth modeled @waller_applied_2004.

Our model of interspecific competition relies on a spatial definition of who the competitor trees are for focal trees of interest: all trees within a distance `comp_dist` of a focal tree are considered its competitors (assuming the same units as the `gx` and `gy` location variables). In our case we set this value at 7.5m, a value informed by @canham_neighborhood_2004 @uriarte_spatially_2004 @canham_neighborhood_2006. Using this value along with a manually constructed `sf` object representation of the study region's boundary, we apply the `add_buffer_variable()` to our `growth_scbi` data frame to add a `buffer` boolean variable: all trees who have `buffer` set to `FALSE` will be our focal trees whose growth will be modeled, whereas those with `TRUE` will only be considered as competitor trees whose growth will not.

@@

The second element of spatial information are blocks corresponding to folds of a spatial cross-validation algorithm used to estimate out-of-sample model error. Conventional cross-validation algorithms assign observations to folds by randomly resampling individual observations. However, an assumption that the observations are independent underlies any cross-validation algorithm. In the case of forest census data, observations exhibit spatial autocorrelation. We therefore incorporate this spatial dependence into the cross-validation algorithm by randomly resampling spatial blocks of trees @roberts_cross-validation_2017 @pohjankukka_estimating_2017. (DAVE NOTE: I don't want to change somethign I don't understand, but "randomly resampling" seems wrong to me. I don't see anything random about our block structure. Seems to me like "incorpoate this spatial dependence into the cross-validation algorithm with our spatial blocks of trees.")

@@

DAVE NOTE::there are a couple of places where you convert to factor. Earlier with `sp` and then twice in the code block below. ANy reason we can't just do this "under the hood" in the package code? Also any reason with swtiched from undercase with underscores to `SpatialBlock_scbi` in the naming?

@@

We then compare the observed and fitted/predicted growths to compute the root mean squared error (RMSE) of our model fit. DAVE NOTE: I rounded all the RMSEs, in ecology convention to not show WAY more digits than resolution of the measurement.

```{r scbi-model-rmse}
model_rmse <- focal_vs_comp_scbi %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate) %>%
  round(3)
model_rmse
```

@@

These plots give the posterior distributions of parameters from equation 1 (DAVE NOTE:SOrry couldn't figure out how to make equation references work). For many pacakge users they will be of interest because they give insight into the species-specific competitive interactions. Setting `type = "intercepts"` gives posterior distributions for $\beta_{0,j}$ and `type = "dbh_slopes"` for $\beta_{dbh,j}$. These give species specific growth independent of competition. The values of more interest are plotted with `type = 'competition'` which gives the posterior distribution for $\lambda_{j,k}$ species-specific competition coefficients (i.e., the $\lambda$-matrix). Negative values indicate a competitor species which slows the growth of a focal species. Here, for example, we see that tulip trees (`litu`) have a strong negative effect on the growth of conspecifics but relatively little effect on neighbors of the other two species.

@@

model_rmse_shuffle <- focal_vs_comp_scbi %>%
  rmse(truth = growth, estimate = growth_hat_shuffle) %>%
  pull(.estimate) %>%
  round(3)
model_rmse_shuffle
```

The resulting RMSE of `r model_rmse_shuffle` based on the permutation test is `r ifelse(model_rmse_shuffle < model_rmse, "smaller", "larger")` than the earlier RMSE of `r model_rmse`, suggesting that models that do incorporate competitor species identity better fit the data. We conduct an analysis of the full SCBI plot in Section below.

@@

To this end, we use the `foldID` variable defined in Section \ref{spatial-information} whereby all focal trees are assigned to one of `r n_fold` spatially contiguous blocks that act as folds in our cross-validation routine. Figure \ref{fig:scbi-spatial-cross-validation-schematic} presents a schematic illustrating this scheme for fold 1 (bottom-left) as the test set and folds 2, 3, and 4 as the training sets. We fit the model to all focal trees in the training set, apply the model to all focal trees in the test set to compute fitted/predicted values, and compute the RMSE of the observed versus predicted growths. We repeat this procedure 3 more times with each of the three remaining folds acting as the test set and then average all four resulting RMSE's. Furthermore, in order to maintain spatial independence between the test and training set, a fold buffer that extend outwards from the boundary of the test set is computed; all trees falling within this fold buffer are excluded from the training set. DAVE NOTE: I repalced "buffer" with "fold buffer" just to keep clear that this is different than our global buffer. But if you think that distinction is not necessary feel free to switch back!

```{r scbi-spatial-cross-validation-schematic, out.width="66%", fig.cap= "Schematic of spatial cross-validation: Using the k = 1 fold as the test set, assigning each focal tree to training set, test set, and fold buffer.", echo = FALSE}
fold_number <- 1

@@

test_vs_train <- focal_vs_comp_scbi %>% 
  st_as_sf() %>% 
  add_buffer_variable(
    direction = "out", size = comp_dist, region = test_fold_buffer
  ) %>% 
  mutate(
    fold = case_when(
      foldID == fold_number ~ "test",
      foldID != fold_number & buffer ~ "training",
      foldID != fold_number & !buffer ~ "fold buffer"
    ),
    fold = factor(fold, levels = c("test", "fold buffer", "training"))
  )

@@

```{r scbi-spatial-cv}
focal_vs_comp_scbi <- focal_vs_comp_scbi %>%
  run_cv(comp_dist = comp_dist, cv_grid = blocks_scbi)
```
```{r}
model_rmse_cv <- focal_vs_comp_scbi %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate) %>%
  round(3)
model_rmse_cv
```
